IALE 2015 Analysis

### 2016-01-26 iale2015_sem.Rmd
Edit to knit reproducibly.

```{r load data}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")
# Vegetation data
veg_us_2005 <- read.csv("analysis/data/understory_veg_2005.csv")
veg_us_2011 <- read.csv("analysis/data/understory_veg_2011.csv")
# Untagged dbh data
untagged_dbh <- read.csv("analysis/data/untagged_dbh_summary.csv")
canopy_cover <- read.csv("analysis/data/plot_canopy_cover_dens.csv")

# Physical environment: topography
library(rgdal)
plots_shp <- readOGR("analysis/data/plot_gis/", "soco_plots_metric")
summary(plots_shp)
plot(plots_shp, axes=T)
plots_env <- plots_shp@data

# Temperature: rainy season hour summaries
library(plyr); library(dplyr)
plots_temps <- read.csv("analysis/data/rainy_season_temps.csv")
summary(plots_temps)
plots_temps <- select(plots_temps, -X)
```

Join the plot level environmental data (geographic/topographic & temperatures)
```{r join plot data, message=FALSE}
plots_env <- left_join(plots_env, plots_temps, by = c("PLOT_ID"="plotid"))
summary(plots_env)
str(plots_env)
plots_env <- plots_env %>% select(-cat) %>% rename(easting = X, northing = Y, plotid = PLOT_ID)
```

Investigating `NAs` in the `plots_env` data set
```{r investigate NAs}
summary(plots_env)
filter(plots_env, is.na(sample_year))

@@

Examine the bay laurel data at individual and plot levels
```{r examine umca data}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.
summary(bay_laurel)
# This has stem-level data on bay laurel.
```

I can use this to calculate plot level leaf count variables (sum, average) for each plot and year
```{r calculate plot level leaf counts}
plots_umca <- left_join(plots_umca, bay_laurel %>% filter(status == "Alive") %>% 
      group_by(plot, year) %>% 
      summarise(
            avg_lfct = mean(slc, na.rm = T),
            tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(plots_umca)
```

```{r examine lfct NAs}
filter(plots_umca, is.na(avg_lfct))
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm removing it manually from the `plots_umca` data frame
plots_umca <- plots_umca %>% filter(avg_lfct != "NaN")
summary(plots_umca)
```

@@

```{r join env and umca data}
summary(plots_env)
plots_env$avg_psi <- rowMeans(cbind(plots_env$psiSpr, plots_env$psiSum, plots_env$psiWin))
summary(plots_umca)
# plot == plotid, year == sample_year
plots_umca <- left_join(select(plots_umca, -species), plots_env, by = c("plot" = "plotid", "year" = "sample_year"))
summary(plots_umca)
filter(plots_umca, is.na(easting))
# NAs are due to no microclimate data for year 2003, and abandoned plots BUSH01 & SUGAR02
```

Now I also want to calculate some plot-level community variables from the vegetation data. First, understory richness from the vegetation transects.
```{r calculate understory richness}
summary(veg_us_2005)
veg_us_2005 <- as.tbl(select(veg_us_2005, -X))
summary(veg_us_2011)
veg_us_2011 <- as.tbl(select(veg_us_2011, -X))
veg_us_2011
veg_us_2005

@@

```{r calculate overstory richness}
summary(untagged_dbh)
untagged_dbh <- as.tbl(untagged_dbh %>% select(-X) %>% rename(plotid = PlotID, species = Species))
untagged_dbh
unique(untagged_dbh$year)
untagged_dbh$year[untagged_dbh$year < 2014] <- 2005 # Create two most complete groups
untagged_dbh$plotid <- tolower(untagged_dbh$plotid)

@@

```{r join richness and umca data}
plots_umca
richness
plots_umca <- rename(plots_umca, plotid = plot)
plots_umca$plotid <- tolower(plots_umca$plotid)

@@

Make a pairs plot of the continuous variables for UMCA plots.
```{r umca sod pairs plot}
# Helper functions for `pairs`
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}

@@

```{r join umca and qusp data}
plots_umca
plot_qusp # right now this is broken down by oak species
summary(plot_qusp)
oak_sod <- plot_qusp %>% group_by(plot, year) %>%
      summarise(uninf_oak_ct = sum(uninfected_oak_ct), inf_oak_ct = sum(infected_oak_ct), tot_oak_ct = sum(c(uninfected_oak_ct, infected_oak_ct)))
oak_sod
oak_sod <- rename(oak_sod, plotid = plot)
oak_sod$plotid <- tolower(oak_sod$plotid)

@@

```{r oak_sod pairs plot}
class(oak_sod)
oak_sod <- ungroup(oak_sod)
pairs(select(oak_sod, -plotid, -year, -starts_with("psi"), -tagged_rich, -untagged_rich), lower.panel = panel.cor, diag.panel = panel.hist, main = "Oak_SOD Data")
```

@@

```{r assess conditional independence claims 2005}
bu_oak_path1 # 15 independence claims means 15 models
library(nlme)
library(lme4)

@@

```{r estimate path coefficients 2005}
# f0 <- fitted(mlme1, level = 0)# Values from population model
# f1 <- fitted(mlme1, level = 1)# Within beach fitted values

@@

```{r assess conditional independence 2014}
f1 <- lme(avg_psi ~ d2c, data = oak_2014, random = ~1|plotid)
summary(f1) # p = 0.591
f2 <- lme(avg_psi ~ elevation, data = oak_2014, random = ~1|plotid)
summary(f2) # p = 0.2171
f3 <- lme(avg_psi ~ tmi + elevation, data = oak_2014, random = ~1|plotid)
summary(f3) # tmi; p = 0.1889
f4 <- lm(os_rich ~ avg_psi + tmi, data = oak_2014)
# f4 <- lme(os_rich ~ avg_psi + tmi, data = oak_2014, random = ~1|plotid)
summary(f4) # avg_psi; p = 0.156
f5 <- lm(log1p(tot_lfct) ~ scale(avg_psi) + scale(d2c) + scale(hrs_14_20) + scale(os_rich), data = oak_2014)
# f5 <- lme(log1p(tot_lfct) ~ scale(avg_psi) + scale(d2c) + scale(hrs_14_20) + scale(os_rich), data = oak_2014, random = ~1|plotid)
# f5 <- glmer(tot_lfct ~ scale(avg_psi) + scale(d2c) + scale(hrs_14_20) + scale(os_rich) + (1|plotid), data = oak_2014, family = "poisson")
summary(f5) # avg_psi; p = 0.265

@@

```{r recalculate c-statistic}
pvalues <- c(
      summary(f1)$tTable[2,5], summary(f2)$tTable[2,5], summary(f3)$tTable[2,5],
      summary(f4)$coefficients[2,4], summary(f5)$coefficients[2,4], 
             # summary(f6)$coefficients[2,4], 
      summary(f7)$coefficients[2,4], summary(f8)$coefficients[2,4],
      summary(f9)$coefficients[2,4], summary(f10)$coefficients[2,4],
      # summary(f11)$coefficients[2,4], 
      # summary(f12)$coefficients[2,4], 
      summary(f13)$coefficients[2,4], summary(f14)$coefficients[2,4], summary(f15)$coefficients[2,4])

@@

```{r estimate path coefficients 2014}
# f0 <- fitted(mlme1, level = 0)# Values from population model
# f1 <- fitted(mlme1, level = 1)# Within beach fitted values

@@

# fit4_2014 <- lm((inf_oak_ct/(inf_oak_ct+uninf_oak_ct)) ~ scale(tot_lfct) + scale(os_rich) + scale(hrs_14_20) + scale(d2c) + scale(elevation) + scale(avg_psi), data = oak_2014)
par(mfrow = c(2,2))
plot(fit4_2014)

@@

# fit4_2014 <- glmer(cbind(inf_oak_ct, uninf_oak_ct) ~ scale(tot_lfct) + scale(os_rich) + scale(hrs_14_20) + scale(d2c) + (1|plotid), data = oak_2014, family = "binomial") # best model
summary(fit4_2014)

@@

```{r calculate importance value}
unique(untagged_dbh$species)
untagged_dbh
unique(stems$species)

>>>>

```{r load data, echo=F, message=F}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")
# Vegetation data
veg_us_2005 <- read.csv("analysis/data/understory_veg_2005.csv")
veg_us_2011 <- read.csv("analysis/data/understory_veg_2011.csv")
# Untagged dbh data
untagged_dbh <- read.csv("analysis/data/untagged_dbh_summary.csv")
canopy_cover <- read.csv("analysis/data/plot_canopy_cover_dens.csv")

# Physical environment: topography
library(rgdal)
plots_shp <- readOGR("analysis/data/plot_gis/", "soco_plots_metric", verbose = F)
summary(plots_shp)
plot(plots_shp, axes=T)
plots_env <- plots_shp@data

# Temperature: rainy season hour summaries
library(plyr); library(dplyr)
plots_temps <- read.csv("analysis/data/rs_temps_data.csv")
summary(plots_temps)
# plots_temps <- select(plots_temps, -X)
```

Join the plot level environmental data (geographic/topographic & temperatures)
```{r join plot data, message=FALSE, echo=F}
plots_env <- left_join(plots_env, plots_temps, by = c("PLOT_ID"="plotid"))
summary(plots_env)
str(plots_env)
plots_env <- plots_env %>% select(-cat) %>% rename(easting = X, northing = Y, plotid = PLOT_ID)
```

Investigating `NAs` in the `plots_env` data set
```{r investigate NAs, echo=F}
summary(plots_env)
filter(plots_env, is.na(sample_year))

@@

Examine the bay laurel data at individual and plot levels
```{r examine umca data, echo=F}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.
summary(bay_laurel)
# This has stem-level data on bay laurel.
```

I can use this to calculate plot level leaf count variables (sum, average) for each plot and year
```{r calculate plot level leaf counts, echo=F}
plots_umca <- left_join(plots_umca, bay_laurel %>% filter(status == "Alive") %>% 
      group_by(plot, year) %>% 
      summarise(
            avg_lfct = mean(slc, na.rm = T),
            tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(plots_umca)
```

```{r examine lfct NAs, echo=F, results='hide'}
filter(plots_umca, is.na(avg_lfct))
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm removing it manually from the `plots_umca` data frame
plots_umca <- plots_umca %>% filter(avg_lfct != "NaN")
summary(plots_umca)
```

@@

```{r join env and umca data, echo=F, results='hide'}
summary(plots_env)
plots_env$avg_psi <- rowMeans(cbind(plots_env$psiSpr, plots_env$psiSum, plots_env$psiWin))
summary(plots_umca)
# plot == plotid, year == sample_year
plots_umca <- left_join(select(plots_umca, -species), plots_env, by = c("plot" = "plotid", "year" = "sample_year"))
summary(plots_umca)
filter(plots_umca, is.na(easting))
# NAs are due to no microclimate data for year 2003, and abandoned plots BUSH01 & SUGAR02
```

Now I also want to calculate some plot-level community variables from the vegetation data. First, understory richness from the vegetation transects.
```{r calculate understory richness, echo=F, results='hide'}
summary(veg_us_2005)
veg_us_2005 <- as.tbl(select(veg_us_2005, -X))
summary(veg_us_2011)
veg_us_2011 <- as.tbl(select(veg_us_2011, -X))
veg_us_2011
veg_us_2005

@@

```{r calculate overstory richness, echo=F, results='hide'}
summary(untagged_dbh)
untagged_dbh <- as.tbl(untagged_dbh %>% select(-X) %>% rename(plotid = PlotID, species = Species))
untagged_dbh
unique(untagged_dbh$year)
untagged_dbh$year[untagged_dbh$year < 2014] <- 2005 # Create two most complete groups
untagged_dbh$plotid <- tolower(untagged_dbh$plotid)

@@

```{r join richness and umca data, echo=F, results='hide'}
plots_umca
richness
plots_umca <- rename(plots_umca, plotid = plot)
plots_umca$plotid <- tolower(plots_umca$plotid)

@@

Make a pairs plot of the continuous variables for UMCA plots.
```{r umca sod pairs plot, echo=F, results='hide'}
# Helper functions for `pairs`
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}

@@

```{r join umca and qusp data, echo=F, results='hide'}
plots_umca
plot_qusp # right now this is broken down by oak species
summary(plot_qusp)
oak_sod <- plot_qusp %>% group_by(plot, year) %>%
      summarise(uninf_oak_ct = sum(uninfected_oak_ct), inf_oak_ct = sum(infected_oak_ct), tot_oak_ct = sum(c(uninfected_oak_ct, infected_oak_ct)))
oak_sod
oak_sod <- rename(oak_sod, plotid = plot)
oak_sod$plotid <- tolower(oak_sod$plotid)

@@

```{r oak_sod pairs plot, echo=F}
class(oak_sod)
oak_sod <- ungroup(oak_sod)
pairs(select(oak_sod, -plotid, -year, -starts_with("psi"), -tagged_rich, -untagged_rich), lower.panel = panel.cor, diag.panel = panel.hist, main = "Oak_SOD Data")
```

@@

```{r assess conditional independence claims 2005, message=F}
bu_oak_path1 # 15 independence claims means 15 models
library(nlme); library(lme4); library(MuMIn)

x.quasibinomial <- function(...) {
    res <- quasibinomial(...)
    res$aic <- binomial(...)$aic
    res
}

x.quasipoisson <- function(...) {
    res <- quasipoisson(...)
    res$aic <- poisson(...)$aic
    res
}

@@

```{r estimate path coefficients 2005, message=F}
# f0 <- fitted(mlme1, level = 0)# Values from population model
# f1 <- fitted(mlme1, level = 1)# Within beach fitted values

@@

```{r assess conditional independence 2014, message=F}
f1 <- lme(avg_psi ~ d2c, data = oak_2014, random = ~1|plotid)
summary(f1) # p = 0.591
f2 <- lme(avg_psi ~ elevation, data = oak_2014, random = ~1|plotid)
summary(f2) # p = 0.2171
f3 <- lme(avg_psi ~ tmi + elevation, data = oak_2014, random = ~1|plotid)
summary(f3) # tmi; p = 0.1889
f4 <- lm(os_rich ~ avg_psi + tmi, data = oak_2014)
# f4 <- lme(os_rich ~ avg_psi + tmi, data = oak_2014, random = ~1|plotid)
summary(f4) # avg_psi; p = 0.156
f5 <- lm(log1p(tot_lfct) ~ scale(avg_psi) + scale(d2c) + scale(hrs_14_20) + scale(os_rich), data = oak_2014)
# f5 <- lme(log1p(tot_lfct) ~ scale(avg_psi) + scale(d2c) + scale(hrs_14_20) + scale(os_rich), data = oak_2014, random = ~1|plotid)
# f5 <- glmer(tot_lfct ~ scale(avg_psi) + scale(d2c) + scale(hrs_14_20) + scale(os_rich) + (1|plotid), data = oak_2014, family = "poisson")
summary(f5) # avg_psi; p = 0.265

@@

```{r recalculate c-statistic, echo=F}
pvalues <- c(
      summary(f1)$tTable[2,5], summary(f2)$tTable[2,5], summary(f3)$tTable[2,5],
      summary(f4)$coefficients[2,4], summary(f5)$coefficients[2,4], 
             # summary(f6)$coefficients[2,4], 
      summary(f7)$coefficients[2,4], summary(f8)$coefficients[2,4],
      summary(f9)$coefficients[2,4], summary(f10)$coefficients[2,4],
      # summary(f11)$coefficients[2,4], 
      # summary(f12)$coefficients[2,4], 
      summary(f13)$coefficients[2,4], summary(f14)$coefficients[2,4], summary(f15)$coefficients[2,4])

@@

```{r estimate path coefficients 2014, message=F}
# f0 <- fitted(mlme1, level = 0)# Values from population model
# f1 <- fitted(mlme1, level = 1)# Within beach fitted values

@@

# fit4_2014 <- lm((inf_oak_ct/(inf_oak_ct+uninf_oak_ct)) ~ scale(tot_lfct) + scale(os_rich) + scale(hrs_14_20) + scale(d2c) + scale(elevation) + scale(avg_psi), data = oak_2014)

@@

# fit4_2014 <- glmer(cbind(inf_oak_ct, uninf_oak_ct) ~ scale(tot_lfct) + scale(os_rich) + scale(hrs_14_20) + scale(d2c) + (1|plotid), data = oak_2014, family = "binomial") # best model
summary(fit4_2014)
par(mfrow = c(2,2))
plot(fit4_2014)

@@

```{r calculate importance value, eval=F, echo=F}
unique(untagged_dbh$species)
untagged_dbh
unique(stems$species)
